Semantic integration of physiology phenotypes: an application to the cellular phenotype ontology

### 2011-12-01 trunk/paper/drinking.tex

\title{Process-based phenotypes: an application to the ontology of
  cell physiology}

@@

\address{$^{1}$Department of Genetics, University of Cambridge,
  Downing Street, Cambridge, Cambridge CB2 3EH, UK\\
  $^{2}$Department of Biochemistry; University of Cambridge, 80 Tennis
  Court Road, Cambridge CB2 1GA, UK\\
  $^{3}$Institute for Medical Informatics, Statistics and
  Epidemiology, University of Leipzig, Haertelstrasse 16-18, 04107
  Leipzig, Germany\\
  $^{4}$EBI, Wellcome Trust Genome Campus, Hinxton, Cambridge,
  Cambridge CB10 1SD, UK
}

@@

\begin{abstract}

  \url{http://cell-phenotype.googlecode.com}.
\end{abstract}

@@

\section{Materials and Methods}
\subsection{Formal ontology}
The ontology as an approach to semantic standardisation was proposed
more than a decade ago and since then has become the dominant
methodology used to semantically categorise phenodeviance.  The
biomedical research community has invested considerable amount of
effort and resources in the development and establishment of
ontologies that are becoming increasingly successful as information
management and integration tools in many disparate scientific fields
allowing interoperability and semantic information processing between
diverse biomedical resources and domains.

@@

When using the GO and PATO ontologies from 30 Nov 2011, CPO contains
125,466 classes of which 79,236 are explicitly defined.  The ELK
reasoner \cite{Kazakov2011} is able to perform a classification of the
ontology in under 10 seconds. We make the ontology and the source code
that is used to generate it freely available on
\url{http://cell-phenotype.googlecode.com}.

@@

A further domain that will greatly benefit from the CPO is {\em
  systems microscopy}, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
\cite{Lock2010}.  If we consider some of the studies, which have been
published in this field in the last few years \cite{Neumann2010,
  Schmitz2010, Fuchs2010}, the need for CPO becomes evident.  In the
three studies illustrated in Table \ref{tbl:studies}, live-cell
imaging assays and RNAi knockdown were used to generate phenotypic
profiles that quantify the cellular response to a given siRNA thus
allowing identification of hundreds of genes involved in diverse
biological functions including cell division, migration and survival.
In each study, several phenotypes were detected and described by the
authors without the use of ontologies (see Table \ref{tbl:studies},
making the integration between datasets extremely difficult. For
example, it is evident that cell division phenotypes were observed in
all three datasets (e.g. {\em Mitotic delay/arrest} and {\em Prolonged
  mitotic exit} or {\em Methaphase delay} and {\em Methaphase cells})
but the overlap between such phenotypes is unclear.

@@

\section{Acknowledgements}
Funding for RH is provided by the European Commission's 7th
Framework Programme, RICORDO project, grant number 248502. MAH is
supported by Wellcome Trust, grant WT090548MA. HH is supported by the
Institute for Medical Informatics, Statistics and Epidemiology,
University of Leipzig. GVG is funded by BBSRC grant BBG0043581.

>>>>

\title{Semantic integration of physiology phenotypes: an application
  to the cellular phenotype ontology}

@@

\address{$^{1}$Department of Genetics, University of Cambridge,
  Downing Street, Cambridge, Cambridge CB2 3EH, UK\\
  $^{2}$Department of Biochemistry; University of Cambridge, 80 Tennis
  Court Road, Cambridge CB2 1GA, UK\\
  $^{3}$Institute for Medical Informatics, Statistics and
  Epidemiology, University of Leipzig, Haertelstrasse 16-18, 04107
  Leipzig, Germany\\
  $^{4}$EBI, Wellcome Trust Genome Campus, Hinxton, Cambridge,
  Cambridge CB10 1SD, UK}

@@

\begin{abstract}
  The systematic observation of phenotypes has become a crucial tool
  of functional genomics, and several large international projects are
  currently underway to identify and characterize the phenotypes that
  are associated with genotypes in several species. To integrate
  phenotype descriptions within and across species, phenotype
  ontologies have been developed. Applying ontologies to unify
  phenotype descriptions in the domain of physiology has been a
  particular challenge due to the high complexity of the underlying
  domain.
  Here, we present the outline of a theory for an ontology of
  physiology-related phenotypes. We provide a formal description of
  process attributes and relate them to the attributes of their parts
  and participants. We apply our theory to create the Cellular
  Phenotype Ontology (CPO). The CPO is an ontology of morphological
  and physiological phenotypic abnormalities of cells, cell components
  and cellular processes. Its prime application is the unification of
  cellular phenotype descriptions across species by providing terms
  and uniform definition patterns. The CPO can further be used for the
  annotation of observed abnormalities in domains, such as systems
  microscopy, in which cellular abnormalities are observed and for
  which no phenotype ontology has been created.  The CPO and
  associated source code is available on
  \url{http://cell-phenotype.googlecode.com}.
\end{abstract}

@@

\section{Materials and Methods}
\subsection{Formal ontology}
The ontology as an approach to semantic standardisation was proposed
more than a decade ago and since then has become the dominant
methodology used to semantically categorise phenodeviance.  The
biomedical research community has invested considerable effort and
resources in the development and establishment of ontologies that are
becoming increasingly successful as information management and
integration tools in many disparate scientific fields allowing
interoperability and semantic information processing between diverse
biomedical resources and domains.

@@

As of November 2011, CPO contains 125,466 classes of which 79,236 are
explicitly defined.  The ELK reasoner \cite{Kazakov2011} is able to
perform a classification of the ontology in under 10 seconds. We make
the ontology and the source code that is used to generate it freely
available on \url{http://cell-phenotype.googlecode.com}.

@@

A further domain that will greatly benefit from the CPO is {\em
  systems microscopy}, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
\cite{Lock2010}.  If we consider some of the studies, which have been
published in this field in the last few years \cite{Neumann2010,
  Schmitz2010, Fuchs2010}, the need for CPO becomes evident.  In the
three studies illustrated in Table \ref{tbl:studies}, live-cell
imaging assays and RNAi knockdown were used to generate phenotypic
profiles that quantify the cellular response to a given siRNA thus
allowing identification of hundreds of genes involved in diverse
biological functions including cell division, migration and survival.
In each study, several phenotypes were detected and described by the
authors without the use of ontologies (see Table \ref{tbl:studies}),
making the integration between datasets extremely difficult. For
example, it is evident that cell division phenotypes were observed in
all three datasets (e.g. {\em Mitotic delay/arrest} and {\em Prolonged
  mitotic exit} or {\em Methaphase delay} and {\em Methaphase cells})
but the overlap between such phenotypes is unclear.

@@

\section{Acknowledgements}
Funding for RH is provided by the European Commission's 7th Framework
Programme, RICORDO project, grant number 248502. MAH is supported by
Wellcome Trust, grant WT090548MA. HH is supported by the Institute for
Medical Informatics, Statistics and Epidemiology, University of
Leipzig.  Funding for GR is provided by the European Union's Seventh
Framework Programme (FP7/2007-2013) under grant agreement n°258068;
EU-FP7-Systems Microscopy NoE. GVG is funded by NIH grant number R01
HG004838-02.
